Policing for Profit and the Undermining of Democracy\thanks{The authors thank XX for their feedback and support. All errors are our responsibility.}

### 2021-01-20 markdown/fees_fines_to.Rmd
keep going

Fines and fees are increasingly recognized as a form of racist revenue extraction connected to marginalized communities’ alienation from government. After Michael Brown was killed by the Ferguson Police Department in 2014, a US Department of Justice investigation into the city’s police and courts demonstrated that the municipality was engaged in a practice that advocates now refer to as “policing for profit.” The city’s reliance on fines and fees to fund government functions grew from 13 to 23 percent of the total budget between fiscal years 2012 and 2015 [@UnitedStatesDepartmentofJusticeCivilRightsDivision2015]. Although a body of literature developed over the past decade indicates that criminalization chills political participation, no work has directly investigated the relationship between fines and fees revenue collection and voter turnout. This project intends to fill that gap.

It wasn’t just a Ferguson problem, or even a Missouri problem. American cities’ reliance on fines and fees revenue increased significantly following the 2008 recession—as local tax revenues dropped and tax increases became less politically viable, jurisdictions increased the amounts of fines and fees and imposed them more frequently in order to fund government services [@Singla2020; @Mughan2020; @Kirk2020; @Martin2018].

@@

Recent work from scholars in sociology and political science, however, indicates that many more voters are *indirectly* disenfranchised due to over-incarceration and the politically chilling effects of the carceral state. Policing and incarceration shape political socialization: personal contact with police or incarceration is associated with a reduced likelihood of political participation, and that alienating effect extends even to people who have incarcerated family members or who simply reside in affected communities [@Lee2014; @Weaver2010; @Morris2020]. This relationship may be more complicated, however. In New York City, concentrated policing was associated with reduced turnout in Congressional elections and increased turnout in the 2008 Presidential election as well as the local mayoral election in 2013 [@Laniyonu2019]. It may also be the case that proximal contact differs from personal contact with the criminal legal system—whereas direct experience with criminalization (and incarceration in particular) reduces the likelihood of voting, proximal or indirect experience produces more mixed results. Hannah Walker suggests that police contact that does not lead to a criminal conviction may even mobilize voters—but her survey questions specifically excluded “minor traffic stops,” a primary site of police ticketing [@Walker2014].

However, one of the most routine interactions Americans have with the police has gone largely unstudied: namely, ticketing for low-level offenses. Police ticketing practices may discourage potential voters without directly disenfranchising them, and ticketing affects more Americans than legal disenfranchisement. In 2018, more than 24 million people experienced police contact in the context of a traffic stop; by contrast, about 5.17 million people are disenfranchised as a result of a felony conviction [@Harrell2020; @Uggen2020].

Ticketing practices may be related to distrust of government and reduced political participation. In a survey of residents in three Georgia cities that rely heavily on fines and fees revenue, residents who were ticketed reported lower levels of trust in police and government more broadly [@CarpenterII2019]. Lerman and Weaver have argued that residents’ perceptions of law enforcement may affect their perceptions of government as a whole. As one man told them, “I feel like if I contact a senator or governor, they’ll probably want to put me in jail and leave me as a troublemaker... I try to basically stay away from government” [-@Lerman2014]. In their study of descriptive representation, Sances and You [-@Sances2017] explicitly suggested that “fines may make descriptive representation less likely by depressing minority turnout,” and called for further research regarding voter turnout and the imposition of fines. In this paper, we clarify that relationship.

@@

Unfortunately, the COG data does not include voter turnout, and turnout data is not generally available at this level. Therefore, in order to determine municipality-level turnout, we employ a national, geocoded voter file. This voter file is provided by L2 Political, and includes a host of information about each voter, including age, party affiliation, racial estimates, and historical turnout. By using the latitude and longitude of every registered voter in the country and shapefiles provided by the Census Bureau, we map each voter to their home municipality using the same definitions as the COG.^[Although the registered voter file includes the name of the city in which the voter is registered, these do not always map neatly to the municipalities defined in the COG data. The use of these shapefiles ensures an apples-to-apples comparison.] We then aggregate these individual-level records up to the municipality level to determine municipal turnout, measured as a share of registered voters. The full study will include other turnout denominators to avoid biased estimates arising from differential registration rates.

@@

### Results {-}

Figure \ref{fig:marg1} makes clear that there is a relationship between the fees and fines collected by a municipality and its turnout. The left hand panel shows the simple bivariate relationship between the two quantities. In the right hand panel, we present the marginal effects plot in lieu of a regression table due to space constraints. The second panel shows that, even after controlling for a municipality’s racial composition, median income, collegiate education, age, and state fixed effects, there is a statistically significant, negative relationship between fees and fines and turnout in 2018. In these models, the dependent variable is municipal turnout in 2018 and the independent variable of interest is the natural log of the dollars in fees and fines per resident plus 1.

@@

All else equal, a 10 percent increase in the dollars of fees and fines raised per resident is associated with a reduction in turnout of 0.06 percentage points, and the relationship is significant at the 99 percent confidence level. Although this may seem small, Figure \ref{fig:marg1} shows that the range of fees and fines raised is large. The figure marks the 10th and 90th percentiles for revenue collected in fees and fines, and predicted turnout drops by 1.8 percentage points over this range.

Figure \ref{fig:marg2} presents the results of preliminary analyses investigating the relationship between fees and fines collection and the racial turnout gap.^[Because we are interested in the racial turnout gap, we include only cities where the population is less than 90 percent non-Hispanic white.] Because the bivariate regression is not statistically significant and the *R^2^* is very small (<0.001) we refrain from including the line of best fit in the first panel. Our initial analysis indicates, however, that the relationship between fees and fines collection and the racial turnout gap in a municipality is moderated by its racial composition. The second panel of Figure \ref{fig:marg2} indicates the racial turnout gap is much larger in minority-white cities that collect no fees and fines than in whiter cities that also do not collect fees and fines. As fees and fines collections increase in less-white cities the turnout gap drops, though we see no such pattern in municipalities that are predominantly white. As in Figure \ref{fig:marg1}, we mark the 10th and 90th percentiles.

```{r, fig.show="hold", out.width="50%", echo = FALSE, message = F, warning = F, fig.cap = "\\label{fig:marg2}Dollars per Resident and 2018 Turnout Gap", fig.pos = "H"}
p3 <- readRDS("./temp/simple_scat_gap.rds")
load("./temp/mef_gap.rdata")
print(p3)
print(p4)
```

@@

The significant negative association in our analysis provides some support for our hypothesis: though our preliminary analyses cannot establish causality, they are consistent with a causal story that links fines and fees practices with lower political participation among those who are subjected to them. On the other hand, officials in municipalities with lower turnout may feel more free to rely on fees and fines for revenue collection because they have less fear of electoral repercussions. Our full study will explore these causal relationships directly by supplementing our initial analysis with an analysis of voter turnout and revenue reliance change over time.

That the racial turnout gap is more sensitive to fees and fines collection in municipalities with a smaller white share of the population is somewhat surprising. This could reflect racially disparate policing in cities with large nonwhite populations. It is possible that cities with many nonwhite residents and *low* fees and fines collections disproportionately penalize nonwhite residents; as fees and fines collections increase, however, a larger share of the population - and, thus, more white voters - may be swept into these programs, thereby *reducing* racial disparities and consequently discrepancies in the racial turnout gap.

>>>>

Fines and fees are increasingly recognized as a form of racist revenue extraction connected to marginalized communities’ alienation from government. After Michael Brown was killed by the Ferguson Police Department in 2014, a US Department of Justice investigation into the city’s police and courts demonstrated that the municipality was engaged in a practice that advocates now refer to as “policing for profit.” The city’s reliance on fines and fees to fund government functions grew from 13 to 23 percent of the total budget between fiscal years 2012 and 2015 [@UnitedStatesDepartmentofJusticeCivilRightsDivision2015].

It wasn’t just a Ferguson problem, or even a Missouri problem. American cities’ reliance on fines and fees revenue increased significantly following the 2008 recession—as local tax revenues dropped and tax increases became less politically viable, jurisdictions increased the amounts of fines and fees and imposed them more frequently in order to fund government services [@Singla2020; @Mughan2020; @Kirk2020; @Martin2018]. Although a body of literature developed over the past decade indicates that criminalization chills political participation, no work has directly investigated the relationship between fines and fees revenue collection and voter turnout. This project intends to fill that gap.

@@

Recent work from scholars in sociology and political science, however, indicates that many more voters are also *indirectly* disenfranchised due to over-incarceration and the politically chilling effects of the carceral state. Policing and incarceration shape political socialization: personal contact with police or incarceration is associated with a reduced likelihood of political participation, and that alienating effect extends even to people who have incarcerated family members or who simply reside in affected communities [@Lee2014; @Weaver2010; @Morris2020]. This relationship is complicated, however. In New York City, concentrated stop-and-frisk activity was associated with *reduced* turnout in Congressional elections, but *increased* turnout in the 2008 presidential election as well as the local mayoral election in 2013 [@Laniyonu2019]. It may also be the case that proximal contact differs from personal contact with the criminal legal system—whereas direct experience with criminalization (and incarceration in particular) reduces the likelihood of voting, proximal or indirect experience produces more mixed results. Hannah Walker suggests that police contact that does not lead to a criminal conviction may even mobilize voters—but her survey questions specifically excluded “minor traffic stops,” a primary site of police ticketing [@Walker2014].

Despite this burgeoning literature, one of the most routine interactions Americans have with the police has gone largely unstudied: namely, ticketing for low-level offenses. Police ticketing practices may discourage potential voters without directly disenfranchising them, and ticketing affects more Americans than legal disenfranchisement. In 2018, more than 24 million people experienced police contact in the context of a traffic stop; by contrast, about 5.17 million people are disenfranchised as a result of a felony conviction [@Harrell2020; @Uggen2020].

Ticketing practices may be related to distrust of government and reduced political participation. In a survey of residents in three Georgia cities that rely heavily on fines and fees revenue, residents who were ticketed reported lower levels of trust in police and government more broadly [@CarpenterII2019]. Lerman and Weaver have argued that residents’ perceptions of law enforcement may affect their perceptions of government as a whole. As one man told them, “I feel like if I contact a senator or governor, they’ll probably want to put me in jail and leave me as a troublemaker... I try to basically stay away from government” [-@Lerman2014, p. 210]. In their study of descriptive representation, Sances and You [-@Sances2017] explicitly suggested that “fines may make descriptive representation less likely by depressing minority turnout,” and called for further research regarding voter turnout and the imposition of fines. In this paper, we clarify that relationship.

@@

Unfortunately, the COG data does not include voter turnout, and turnout data is not generally available at this level. Therefore, in order to determine municipality-level turnout, we employ a national, geocoded voter file. This voter file is provided by L2 Political, and includes a host of information about each voter, including age, party affiliation, racial estimates, and historical turnout. By using the latitude and longitude of every registered voter in the country and shapefiles provided by the Census Bureau, we map each voter to their home municipality using the same definitions as the COG.^[Although the registered voter file includes the name of the city in which the voter is registered, these do not always map neatly to the municipalities defined in the COG data. The use of these shapefiles ensures an apples-to-apples comparison.] We then aggregate these individual-level records up to the municipality level to determine the number of ballots cast in each municipality. TUrnout is calculated by dividing this number by the city’s citizen voting age population (CVAP) from the Census Bureau.

@@

### Preliminary Results {-}

Figure \ref{fig:marg1} makes clear that there is a relationship between the fees and fines collected by a municipality and its turnout. The left-hand panel shows the simple bivariate relationship between the two quantities. In the right-hand panel, we present the marginal effects plot in lieu of a regression table due to space constraints. The second panel shows that, even after controlling for a municipality’s racial composition, median income, collegiate education, age, and state fixed effects, there is a statistically significant, negative relationship between fees and fines and turnout in 2018. In these models, the dependent variable is municipal turnout in 2018 and the independent variable of interest is the natural log of the dollars in fees and fines per resident plus 1.

@@

All else equal, a 1 percent increase in the dollars of fees and fines raised per resident is associated with a reduction in turnout of 0.003 percentage points, and the relationship is significant at the 99 percent confidence level. Although this may seem small, Figure \ref{fig:marg1} shows that the range of fees and fines raised is large. The figure marks the 10th and 90th percentiles for revenue collected in fees and fines, and predicted turnout drops by 0.8 percentage points over this range.

Figure \ref{fig:marg2} presents the results of preliminary analyses investigating the relationship between fees and fines collection and the white-Black turnout gap. Racial turnout figures are calculated by using the number of ballots cast by members of each racial group according to L2 divided by the CVAP estimates for that group in each municipality. Our initial analysis indicates that the white-Black turnout gap in a municipality is strongly associated with the level of fees and fines it imposes. Figure \ref{fig:marg2} shows that the racial turnout gap is much larger in cities where there are fewer fees and fines imposed. Moving from the 10th to 90th percentile of fees and fines per capita is associated with a 9 percentage point drop in the white-Black turnout gap.

```{r, fig.show="hold", out.width="50%", echo = FALSE, message = F, warning = F, fig.cap = "\\label{fig:marg2}Dollars per Resident and 2018 Turnout Gap", fig.pos = "H"}
p5 <- readRDS("./temp/simple_scat_gap_bw.rds")
load("./temp/mef_gap_bw.rdata")
print(p5)
print(p6)
```

@@

The significant negative association in our analysis provides some support for our first hypothesis: though our preliminary analyses cannot establish causality, they are consistent with a causal story that links fines and fees practices with lower political participation among those who are subjected to them. On the other hand, officials in municipalities with lower turnout may feel more free to rely on fees and fines for revenue collection because they have less fear of electoral repercussions. Our full study will explore these causal relationships directly by supplementing our initial analysis with an analysis of voter turnout and revenue reliance change over time.

That the racial turnout gap is smaller in cities with more fees and fines is surprising, and the opposite of what we expected to find. This could reflect racially disparate policing in cities with low fee collection. It is possible that cities with *low* fees and fines collections disproportionately penalize nonwhite residents; as fees and fines collections increase, however, a larger share of the population --- and, thus, more white voters --- may have a negative interaction with a police officer, thereby *reducing* racial disparities and consequently discrepancies in the racial turnout gap.

### 2021-01-20 markdown/fees_fines_to.tex
keep going

\title{Policing for Profit and the Undermining of Democracy\thanks{The authors thank XX for their feedback and support. All errors are our responsibility.}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Fines, Fees, and Turnout}
\author{Jonathan Ben-Menachem\footnote{PhD Student, Columbia University, Department of Sociology (\href{mailto:jbenmenachem@gmail.com}{\nolinkurl{jbenmenachem@gmail.com}})} \and Kevin Morris\footnote{Researcher, Brennan Center for Justice at NYU School of Law (\href{mailto:kevin.morris@nyu.edu}{\nolinkurl{kevin.morris@nyu.edu}})}}
\date{January 13, 2021}

@@

Fines and fees are increasingly recognized as a form of racist revenue extraction connected to marginalized communities' alienation from government. After Michael Brown was killed by the Ferguson Police Department in 2014, a US Department of Justice investigation into the city's police and courts demonstrated that the municipality was engaged in a practice that advocates now refer to as ``policing for profit.'' The city's reliance on fines and fees to fund government functions grew from 13 to 23 percent of the total budget between fiscal years 2012 and 2015 (United States Department of Justice Civil Rights Division \protect\hyperlink{ref-UnitedStatesDepartmentofJusticeCivilRightsDivision2015}{2015}). Although a body of literature developed over the past decade indicates that criminalization chills political participation, no work has directly investigated the relationship between fines and fees revenue collection and voter turnout. This project intends to fill that gap.

It wasn't just a Ferguson problem, or even a Missouri problem. American cities' reliance on fines and fees revenue increased significantly following the 2008 recession---as local tax revenues dropped and tax increases became less politically viable, jurisdictions increased the amounts of fines and fees and imposed them more frequently in order to fund government services (Singla, Kirschner, and Stone \protect\hyperlink{ref-Singla2020}{2020}; Mughan \protect\hyperlink{ref-Mughan2020}{2020}; Kirk, Fernandes, and Friedman \protect\hyperlink{ref-Kirk2020}{2020}; Martin \protect\hyperlink{ref-Martin2018}{2018}).

Fines and fees practices disproportionately target Black Americans: city government reliance on fines and fees revenue is associated with larger Black populations, racial stereotypes affect the imposition of fines and fees, and Black defendants face more severe penalties related to monetary sanctions (United States Commission on Civil Rights \protect\hyperlink{ref-UnitedStatesCommissiononCivilRights2017}{2017}; Harris, Evans, and Beckett \protect\hyperlink{ref-Harris2011}{2011}; Edwards and Harris \protect\hyperlink{ref-Edwards2020}{n.d.}). The relationship between cities' reliance on fines and fees and the proportion of Black residents is significantly diminished when Black communities are represented by at least one Black legislator, however (Sances and You \protect\hyperlink{ref-Sances2017}{2017}). This could suggest that higher reliance on fines and fees results from government officials targeting individuals who hold less political power (Makowsky and Stratmann \protect\hyperlink{ref-Makowsky2009}{2009}).

@@

Recent work from scholars in sociology and political science, however, indicates that many more voters are \emph{indirectly} disenfranchised due to over-incarceration and the politically chilling effects of the carceral state. Policing and incarceration shape political socialization: personal contact with police or incarceration is associated with a reduced likelihood of political participation, and that alienating effect extends even to people who have incarcerated family members or who simply reside in affected communities (Lee, Porter, and Comfort \protect\hyperlink{ref-Lee2014}{2014}; Weaver and Lerman \protect\hyperlink{ref-Weaver2010}{2010}; Morris \protect\hyperlink{ref-Morris2020}{2020}). This relationship may be more complicated, however. In New York City, concentrated policing was associated with reduced turnout in Congressional elections and increased turnout in the 2008 Presidential election as well as the local mayoral election in 2013 (Laniyonu \protect\hyperlink{ref-Laniyonu2019}{2019}). It may also be the case that proximal contact differs from personal contact with the criminal legal system---whereas direct experience with criminalization (and incarceration in particular) reduces the likelihood of voting, proximal or indirect experience produces more mixed results. Hannah Walker suggests that police contact that does not lead to a criminal conviction may even mobilize voters---but her survey questions specifically excluded ``minor traffic stops,'' a primary site of police ticketing (Walker \protect\hyperlink{ref-Walker2014}{2014}).

However, one of the most routine interactions Americans have with the police has gone largely unstudied: namely, ticketing for low-level offenses. Police ticketing practices may discourage potential voters without directly disenfranchising them, and ticketing affects more Americans than legal disenfranchisement. In 2018, more than 24 million people experienced police contact in the context of a traffic stop; by contrast, about 5.17 million people are disenfranchised as a result of a felony conviction (Harrell and Davis \protect\hyperlink{ref-Harrell2020}{2020}; Uggen et al. \protect\hyperlink{ref-Uggen2020}{2020}).

Ticketing practices may be related to distrust of government and reduced political participation. In a survey of residents in three Georgia cities that rely heavily on fines and fees revenue, residents who were ticketed reported lower levels of trust in police and government more broadly (Carpenter II, Sweetland, and McDonald \protect\hyperlink{ref-CarpenterII2019}{2019}). Lerman and Weaver have argued that residents' perceptions of law enforcement may affect their perceptions of government as a whole. As one man told them, ``I feel like if I contact a senator or governor, they'll probably want to put me in jail and leave me as a troublemaker\ldots{} I try to basically stay away from government'' (\protect\hyperlink{ref-Lerman2014}{2014}). In their study of descriptive representation, Sances and You (\protect\hyperlink{ref-Sances2017}{2017}) explicitly suggested that ``fines may make descriptive representation less likely by depressing minority turnout,'' and called for further research regarding voter turnout and the imposition of fines. In this paper, we clarify that relationship.

@@

Unfortunately, the COG data does not include voter turnout, and turnout data is not generally available at this level. Therefore, in order to determine municipality-level turnout, we employ a national, geocoded voter file. This voter file is provided by L2 Political, and includes a host of information about each voter, including age, party affiliation, racial estimates, and historical turnout. By using the latitude and longitude of every registered voter in the country and shapefiles provided by the Census Bureau, we map each voter to their home municipality using the same definitions as the COG.\footnote{Although the registered voter file includes the name of the city in which the voter is registered, these do not always map neatly to the municipalities defined in the COG data. The use of these shapefiles ensures an apples-to-apples comparison.} We then aggregate these individual-level records up to the municipality level to determine municipal turnout, measured as a share of registered voters. The full study will include other turnout denominators to avoid biased estimates arising from differential registration rates.

@@

\hypertarget{results}{%
\subsubsection*{Results}\label{results}}
\addcontentsline{toc}{subsubsection}{Results}

Figure \ref{fig:marg1} makes clear that there is a relationship between the fees and fines collected by a municipality and its turnout. The left hand panel shows the simple bivariate relationship between the two quantities. In the right hand panel, we present the marginal effects plot in lieu of a regression table due to space constraints. The second panel shows that, even after controlling for a municipality's racial composition, median income, collegiate education, age, and state fixed effects, there is a statistically significant, negative relationship between fees and fines and turnout in 2018. In these models, the dependent variable is municipal turnout in 2018 and the independent variable of interest is the natural log of the dollars in fees and fines per resident plus 1.

@@

All else equal, a 10 percent increase in the dollars of fees and fines raised per resident is associated with a reduction in turnout of 0.06 percentage points, and the relationship is significant at the 99 percent confidence level. Although this may seem small, Figure \ref{fig:marg1} shows that the range of fees and fines raised is large. The figure marks the 10th and 90th percentiles for revenue collected in fees and fines, and predicted turnout drops by 1.8 percentage points over this range.

Figure \ref{fig:marg2} presents the results of preliminary analyses investigating the relationship between fees and fines collection and the racial turnout gap.\footnote{Because we are interested in the racial turnout gap, we include only cities where the population is less than 90 percent non-Hispanic white.} Because the bivariate regression is not statistically significant and the \emph{R\textsuperscript{2}} is very small (\textless0.001) we refrain from including the line of best fit in the first panel. Our initial analysis indicates, however, that the relationship between fees and fines collection and the racial turnout gap in a municipality is moderated by its racial composition. The second panel of Figure \ref{fig:marg2} indicates the racial turnout gap is much larger in minority-white cities that collect no fees and fines than in whiter cities that also do not collect fees and fines. As fees and fines collections increase in less-white cities the turnout gap drops, though we see no such pattern in municipalities that are predominantly white. As in Figure \ref{fig:marg1}, we mark the 10th and 90th percentiles.

@@

The significant negative association in our analysis provides some support for our hypothesis: though our preliminary analyses cannot establish causality, they are consistent with a causal story that links fines and fees practices with lower political participation among those who are subjected to them. On the other hand, officials in municipalities with lower turnout may feel more free to rely on fees and fines for revenue collection because they have less fear of electoral repercussions. Our full study will explore these causal relationships directly by supplementing our initial analysis with an analysis of voter turnout and revenue reliance change over time.

That the racial turnout gap is more sensitive to fees and fines collection in municipalities with a smaller white share of the population is somewhat surprising. This could reflect racially disparate policing in cities with large nonwhite populations. It is possible that cities with many nonwhite residents and \emph{low} fees and fines collections disproportionately penalize nonwhite residents; as fees and fines collections increase, however, a larger share of the population - and, thus, more white voters - may be swept into these programs, thereby \emph{reducing} racial disparities and consequently discrepancies in the racial turnout gap.

Our full study will include a more narrow analysis of turnout in jurisdictions that dramatically increased or decreased ticketing over a comparatively short period of time. For instance, a Florida Times-Union investigation revealed that police in four counties disproportionately issued tickets to Black pedestrians (Sanders and Conark \protect\hyperlink{ref-Sanders2017}{2017}). In Tampa, police issued thousands of tickets to Black cyclists between 2012 and 2015---after journalists reported on the problem, the city began to issue fewer than 100 citations each year as of 2018 (Zayas \protect\hyperlink{ref-Zayas2015}{2015}; Frago \protect\hyperlink{ref-Frago2019}{2019}). Systematically conducting such an analysis would likely require collecting and organizing new data, as city-level police ticketing data is inconsistent and not always public information.

@@

\leavevmode\hypertarget{ref-Edwards2020}{}%
Edwards, Frank, and Alexes Harris. n.d. ``An Analysis of Court Imposed Monetary Sanctions in Seattle Municipal Courts, 2000-2017.'' Accessed January 13, 2021. \url{https://doi.org/10.31235/osf.io/ajpqc}.

@@

\leavevmode\hypertarget{ref-Sanders2017}{}%
Sanders, Topher, and Benjamin Conark. 2017. ``Florida Police Issue Hundreds of Bad Pedestrian Tickets Every Year Because They Don't Seem to Know the Law.'' \emph{ProPublica}, December 20, 2017. \url{https://www.propublica.org/article/florida-police-issue-hundreds-of-bad-pedestrian-tickets-every-year-because-they-dont-seem-to-know-the-law?token=CpfrqXaMuR8UnJ5-FryRkVwuKl3C98Ae}.

>>>>

\title{Policing for Profit and the Undermining of Democracy\thanks{The authors thank XX for their feedback and support. All errors are our responsibility.}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Fines, Fees, and Turnout}
\author{Jonathan Ben-Menachem\footnote{PhD Student, Columbia University, Department of Sociology (\href{mailto:jbenmenachem@gmail.com}{\nolinkurl{jbenmenachem@gmail.com}})} \and Kevin Morris\footnote{Researcher, Brennan Center for Justice at NYU School of Law (\href{mailto:kevin.morris@nyu.edu}{\nolinkurl{kevin.morris@nyu.edu}})}}
\date{January 14, 2021}

@@

Fines and fees are increasingly recognized as a form of racist revenue extraction connected to marginalized communities' alienation from government. After Michael Brown was killed by the Ferguson Police Department in 2014, a US Department of Justice investigation into the city's police and courts demonstrated that the municipality was engaged in a practice that advocates now refer to as ``policing for profit.'' The city's reliance on fines and fees to fund government functions grew from 13 to 23 percent of the total budget between fiscal years 2012 and 2015 (United States Department of Justice Civil Rights Division \protect\hyperlink{ref-UnitedStatesDepartmentofJusticeCivilRightsDivision2015}{2015}).

It wasn't just a Ferguson problem, or even a Missouri problem. American cities' reliance on fines and fees revenue increased significantly following the 2008 recession---as local tax revenues dropped and tax increases became less politically viable, jurisdictions increased the amounts of fines and fees and imposed them more frequently in order to fund government services (Singla, Kirschner, and Stone \protect\hyperlink{ref-Singla2020}{2020}; Mughan \protect\hyperlink{ref-Mughan2020}{2020}; Kirk, Fernandes, and Friedman \protect\hyperlink{ref-Kirk2020}{2020}; Martin \protect\hyperlink{ref-Martin2018}{2018}). Although a body of literature developed over the past decade indicates that criminalization chills political participation, no work has directly investigated the relationship between fines and fees revenue collection and voter turnout. This project intends to fill that gap.

Fines and fees practices disproportionately target Black Americans: city government reliance on fines and fees revenue is associated with larger Black populations, racial stereotypes affect the imposition of fines and fees, and Black defendants face more severe penalties related to monetary sanctions (United States Commission on Civil Rights \protect\hyperlink{ref-UnitedStatesCommissiononCivilRights2017}{2017}; Harris, Evans, and Beckett \protect\hyperlink{ref-Harris2011}{2011}; Edwards and Harris \protect\hyperlink{ref-Edwards2020}{2020}). The relationship between cities' reliance on fines and fees and the proportion of Black residents is significantly diminished when Black communities are represented by at least one Black legislator, however (Sances and You \protect\hyperlink{ref-Sances2017}{2017}). This could suggest that higher reliance on fines and fees results from government officials targeting individuals who hold less political power (Makowsky and Stratmann \protect\hyperlink{ref-Makowsky2009}{2009}).

@@

Recent work from scholars in sociology and political science, however, indicates that many more voters are also \emph{indirectly} disenfranchised due to over-incarceration and the politically chilling effects of the carceral state. Policing and incarceration shape political socialization: personal contact with police or incarceration is associated with a reduced likelihood of political participation, and that alienating effect extends even to people who have incarcerated family members or who simply reside in affected communities (Lee, Porter, and Comfort \protect\hyperlink{ref-Lee2014}{2014}; Weaver and Lerman \protect\hyperlink{ref-Weaver2010}{2010}; Morris \protect\hyperlink{ref-Morris2020}{2020}). This relationship is complicated, however. In New York City, concentrated stop-and-frisk activity was associated with \emph{reduced} turnout in Congressional elections, but \emph{increased} turnout in the 2008 presidential election as well as the local mayoral election in 2013 (Laniyonu \protect\hyperlink{ref-Laniyonu2019}{2019}). It may also be the case that proximal contact differs from personal contact with the criminal legal system---whereas direct experience with criminalization (and incarceration in particular) reduces the likelihood of voting, proximal or indirect experience produces more mixed results. Hannah Walker suggests that police contact that does not lead to a criminal conviction may even mobilize voters---but her survey questions specifically excluded ``minor traffic stops,'' a primary site of police ticketing (Walker \protect\hyperlink{ref-Walker2014}{2014}).

Despite this burgeoning literature, one of the most routine interactions Americans have with the police has gone largely unstudied: namely, ticketing for low-level offenses. Police ticketing practices may discourage potential voters without directly disenfranchising them, and ticketing affects more Americans than legal disenfranchisement. In 2018, more than 24 million people experienced police contact in the context of a traffic stop; by contrast, about 5.17 million people are disenfranchised as a result of a felony conviction (Harrell and Davis \protect\hyperlink{ref-Harrell2020}{2020}; Uggen et al. \protect\hyperlink{ref-Uggen2020}{2020}).

Ticketing practices may be related to distrust of government and reduced political participation. In a survey of residents in three Georgia cities that rely heavily on fines and fees revenue, residents who were ticketed reported lower levels of trust in police and government more broadly (Carpenter II, Sweetland, and McDonald \protect\hyperlink{ref-CarpenterII2019}{2019}). Lerman and Weaver have argued that residents' perceptions of law enforcement may affect their perceptions of government as a whole. As one man told them, ``I feel like if I contact a senator or governor, they'll probably want to put me in jail and leave me as a troublemaker\ldots{} I try to basically stay away from government'' (\protect\hyperlink{ref-Lerman2014}{2014}, 210). In their study of descriptive representation, Sances and You (\protect\hyperlink{ref-Sances2017}{2017}) explicitly suggested that ``fines may make descriptive representation less likely by depressing minority turnout,'' and called for further research regarding voter turnout and the imposition of fines. In this paper, we clarify that relationship.

@@

Unfortunately, the COG data does not include voter turnout, and turnout data is not generally available at this level. Therefore, in order to determine municipality-level turnout, we employ a national, geocoded voter file. This voter file is provided by L2 Political, and includes a host of information about each voter, including age, party affiliation, racial estimates, and historical turnout. By using the latitude and longitude of every registered voter in the country and shapefiles provided by the Census Bureau, we map each voter to their home municipality using the same definitions as the COG.\footnote{Although the registered voter file includes the name of the city in which the voter is registered, these do not always map neatly to the municipalities defined in the COG data. The use of these shapefiles ensures an apples-to-apples comparison.} We then aggregate these individual-level records up to the municipality level to determine the number of ballots cast in each municipality. TUrnout is calculated by dividing this number by the city's citizen voting age population (CVAP) from the Census Bureau.

@@

\hypertarget{preliminary-results}{%
\subsubsection*{Preliminary Results}\label{preliminary-results}}
\addcontentsline{toc}{subsubsection}{Preliminary Results}

Figure \ref{fig:marg1} makes clear that there is a relationship between the fees and fines collected by a municipality and its turnout. The left-hand panel shows the simple bivariate relationship between the two quantities. In the right-hand panel, we present the marginal effects plot in lieu of a regression table due to space constraints. The second panel shows that, even after controlling for a municipality's racial composition, median income, collegiate education, age, and state fixed effects, there is a statistically significant, negative relationship between fees and fines and turnout in 2018. In these models, the dependent variable is municipal turnout in 2018 and the independent variable of interest is the natural log of the dollars in fees and fines per resident plus 1.

@@

All else equal, a 1 percent increase in the dollars of fees and fines raised per resident is associated with a reduction in turnout of 0.003 percentage points, and the relationship is significant at the 99 percent confidence level. Although this may seem small, Figure \ref{fig:marg1} shows that the range of fees and fines raised is large. The figure marks the 10th and 90th percentiles for revenue collected in fees and fines, and predicted turnout drops by 0.8 percentage points over this range.

Figure \ref{fig:marg2} presents the results of preliminary analyses investigating the relationship between fees and fines collection and the white-Black turnout gap. Racial turnout figures are calculated by using the number of ballots cast by members of each racial group according to L2 divided by the CVAP estimates for that group in each municipality. Our initial analysis indicates that the white-Black turnout gap in a municipality is strongly associated with the level of fees and fines it imposes. Figure \ref{fig:marg2} shows that the racial turnout gap is much larger in cities where there are fewer fees and fines imposed. Moving from the 10th to 90th percentile of fees and fines per capita is associated with a 9 percentage point drop in the white-Black turnout gap.

@@

The significant negative association in our analysis provides some support for our first hypothesis: though our preliminary analyses cannot establish causality, they are consistent with a causal story that links fines and fees practices with lower political participation among those who are subjected to them. On the other hand, officials in municipalities with lower turnout may feel more free to rely on fees and fines for revenue collection because they have less fear of electoral repercussions. Our full study will explore these causal relationships directly by supplementing our initial analysis with an analysis of voter turnout and revenue reliance change over time.

That the racial turnout gap is smaller in cities with more fees and fines is surprising, and the opposite of what we expected to find. This could reflect racially disparate policing in cities with low fee collection. It is possible that cities with \emph{low} fees and fines collections disproportionately penalize nonwhite residents; as fees and fines collections increase, however, a larger share of the population --- and, thus, more white voters --- may have a negative interaction with a police officer, thereby \emph{reducing} racial disparities and consequently discrepancies in the racial turnout gap.

Our full study will include a more narrow analysis of turnout in jurisdictions that dramatically increased or decreased ticketing over a comparatively short period of time. For instance, a Florida Times-Union investigation revealed that police in four counties disproportionately issued tickets to Black pedestrians (Sanders and Conarck \protect\hyperlink{ref-Sanders2017}{2017}). In Tampa, police issued thousands of tickets to Black cyclists between 2012 and 2015---after journalists reported on the problem, the city began to issue fewer than 100 citations each year as of 2018 (Zayas \protect\hyperlink{ref-Zayas2015}{2015}; Frago \protect\hyperlink{ref-Frago2019}{2019}). Systematically conducting such an analysis would likely require collecting and organizing new data, as city-level police ticketing data is inconsistent and not always public information.

@@

\leavevmode\hypertarget{ref-Edwards2020}{}%
Edwards, Frank, and Alexes Harris. 2020. ``An Analysis of Court Imposed Monetary Sanctions in Seattle Municipal Courts, 2000-2017.'' August. \url{https://doi.org/10.31235/osf.io/ajpqc}.

@@

\leavevmode\hypertarget{ref-Sanders2017}{}%
Sanders, Topher, and Benjamin Conarck. 2017. ``Florida Police Issue Hundreds of Bad Pedestrian Tickets Every Year Because They Don't Seem to Know the Law.'' \emph{ProPublica}, December 20, 2017. \url{https://www.propublica.org/article/florida-police-issue-hundreds-of-bad-pedestrian-tickets-every-year-because-they-dont-seem-to-know-the-law?token=CpfrqXaMuR8UnJ5-FryRkVwuKl3C98Ae}.
